---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Isaac Miller"
date: "2026-03-03"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use th is to make sure warnings don't show upl
---

[deployed html url](https://isaacmiller1113.github.io/workshop-08_winter-2026/code/workshop-code.html)

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschscholzia californica_   | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx( # using xslx function
  here( # relative file path; using here to point where the data is in our repo
    "data",
    "Valliere_etal_EcoApps_Data.xlsx"
  ),
  sheet = "First Harvest" # read in first sheet of excel file; first sheet is the one we wanna look at 
) 

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# create a new object for clean data
drought_exp_clean <- drought_exp |> 
  # cleaning column names
  clean_names() |> 
  # creating a column for new species names
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # creating new column for full  treatment names
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # made species_name and water_treatment a factor and set levels
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting columns of interest 
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: display data frame structure

str(drought_exp_clean)

```

Insert code to display 10 random rows from the data frame here.  


```{r}
#| label: displaying 10 rows

slice_sample(
  drought_exp_clean,
  n = 10
)
```


# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 


```{r}
#| label: SLA and plant mass plot 

ggplot(data = drought_exp_clean,
       aes(x = sla,
           y = total_g)) + 
  geom_point(color = "forestgreen") + 
  labs(x = "Specific leaf area (cm\U00B2/g)",
       y = "Plant mass (g)",
       title = "As SLA increases, total mass also increases") + 
  theme(plot.title.position = "plot")


```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.


```{r}
#| label: comparing mass between water treatments

ggplot(data = drought_exp_clean,
      aes(x = water_treatment,
          y = total_g,
          color = water_treatment)) + 
  geom_jitter(height = 0,
              width = 0.3) + 
  scale_color_manual(values = c(
    "Drought stressed" = ds_col,
    "Well watrered" = ww_col
  )) +
  labs(x = "Treatment",
       y = "Plant Mass (g)",
       title = "Well watered plants have more mass than drought stressed plants") +
  theme(legend.position = "none",
        plot.title.position = "plot")

```


## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass _and_ wrap the long species names using the `label_wrap()` function from the `scales` package.

```{r} 
#| label: plant mass by species

ggplot(data = drought_exp_clean,
       aes(x = reorder(species_name, -total_g), # reorder a factor, 
           y = total_g,
           color = species_name)) + 
  geom_jitter(height = 0,
              width = 0.2) + 
  scale_color_manual(values = c(
    "Grindelia camporum" = gricam_col,
    "Eschscholzia californica" = esccal_col,
    "Stipa pulchra" = stipul_col,
    "Encelia californica" = enccal_col,
    "Salvia leucophylla" = salleu_col,
    "Penstemon centranthifolius" = pencen_col,
    "Lotus scoparius" = lotsco_col
  )) + 
  scale_x_discrete(labels = label_wrap(10)) +
  labs( x = "Species", 
        y = "Plant mass (g)",
        title = "Plant mass differs across species") + 
  theme(legend.position = "none") 

```


# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model

model0 <- lm(
  total_g ~ 1, 
  data = drought_exp_clean
)

# model 1: all predictors

model1 <- lm(
  total_g ~ sla + water_treatment + species_name,
  data = drought_exp_clean
)

# model 2: SLA and water treatment

model2 <- lm( 
  total_g ~ sla + water_treatment,
  data = drought_exp_clean)

# model 3: SLA and species

model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)

# model 4: water treatment and species

model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean
)

# model 5: SLA

model5 <- lm(
  total_g ~ sla, 
  data = drought_exp_clean
)

# model 6: water treatment

model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species

model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)

```

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: model selection

AICc(
  model0, 
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  arrange(AICc) # arrange by descending AICc 

```

`model4` is the best model because it has the lowest AICc, and best describes the variation in plant mass. 
In this particular scenario the best predictors that describe variation in plant mass (g) are water treatment and species, not specific leaf area. 

## Model diagnostics

Insert code to look at the diagnostic plots for the best model.

```{r}
#| label: residual plots for model 4 

par(mfrow = c(2, 2)) # displays plots in 2x2 grid
plot(model4) # plotting model 4 diagnostic plots 


```

Diagnostic plots look good. 

Residuals (standing in for errors) look normally distributed based on QQ plot. 

Residuals look homoscedastic (have constant variance) based on Residuals vs. Fitted plot and Scale-Location plot. 

No outliers influencing model estimates of coefficients.

## Summary

Insert code to display the summary for the best model. 

```{r}
#| label: summary of model 4

summary(model4)

```


**What are the reference levels?** (whatever doesnt show up)

Drought stressed is the reference level for treatment. (has the lowest mass compared to well-watered)

Lotus scoparius is the rference for species (has the lowest mass of all the species)


**What does the `(Intercept)` represent?**  

The intercept is the model prediction for plant mass for Lotus scoparius in the drought stressed treatment. The model predicts that Lotus scoparius mass in the drought stressed treatment will be 0.055 ± 0.025 (SE) g.

**What does `water_treatmentWell watered` represent?**

This is the model prediction for the difference in Lotus scoparius mass between the well watered and drought stressed treatments. The model predicts that this difference is 0.12 ± 0.017 g.

**What does `species_namePenstemon centranthifolius` represent?**  

This is the model prediction for the difference between Penstemon centranthifolius mass in the well watered treatment and Lotus scoparius mass in the well watered treatemnt. The model predicts that this difference is 0.05 ± 0.03 g.

**What does `species_nameEschscholzia californica` represent?**

This is the model prediction for the difference betwen Eschscholzia californica mass in the well watered treatment and Lotus scoparius mass in the well watered treatment. The model predicts that this difference is 0.23 ± 0.03 g.


## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors

**How much variation in the response (total mass, in grams) does this model explain?**

this model explains 72% of the variation in plant mass. 

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

Treatment influences plant mass. 

Species influences plant mass.

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: generating predictions

preds <- ggpredict(
  model4, # model object 
  terms = c("water_treatment", "species_name") # predictors 
) |> 
  # rename columns
  rename(water_treatment = x,
         species_name = group)

# each row represents a model prediction for the mean plant mass for a species in a given water treatment

```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final publication figure
#| fig-height: 6
#| fig-width: 12

ggplot(data = drought_exp_clean,
       aes(x = water_treatment,
           y = total_g,
           color = water_treatment)) + 
  geom_jitter(height = 0,
              width = 0.2,
              alpha = 0.5, # slightly transparent
              shape = 21) + # use open circle
  geom_pointrange(data = preds, 
                  aes(x = water_treatment,
                      y = predicted,
                      ymin = conf.low,
                      ymax = conf.high)) +
  scale_color_manual(values = c(
    "Drought Stressed" = ds_col,
    "Well watered" = ww_col
  )) +
  stat_summary(fun = mean,
               geom = "point",
               color = "red",
               size = 2) +
  scale_x_discrete(labels = label_wrap(10)) + 
  labs(x = "Treatment",
       y = "Plant mass (g)") + 
  facet_wrap(~ species_name,
             nrow = 1) + # everything on the same row
  theme(legend.position = "none")

```

means are in red, model predictions are in black. reason the means are different than predictions is because the predictions take into account the different species and treatments. model accounts for variation in other species. 

# 6. Writing

## Figure caption

**Figure 4. Plant mass (g) differs accross species and treatments.**
Open circles represent individuals in each treatment (red: drought stressed, blue: well watered). Panels depict different species. Opaque points represent model predictions for plant mass, and bars represemnt a 95% confidence intervals of model predictions. Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

Note:

the figure number (Figure 4) and title (Plant mass (g) differs across species and treatments) are in bold
the caption includes a description of the open circles (individual observations) and opaque points (model predictions) and bars (95% CI around predictions) and colors (representing treatments)
data source is copied from the data citation in section 2 (“Clean data”)
## Basic components of test

**predictors**  

treatment and species

**response**  
plant mass (g)

**test**  
linear model (multiple linear regression)

**distribution**  
F 

**parameters (degrees of freedom)**  
7, 62

**test statistic** (f-stat)
27.08

**R^2^**  
0.73

**p-value**   
p < 0.001 

**significance level**  
$\alpha$ = 0.05 

**reference level(s)**  
Drought stressed, and well watered

## Written communication

### Methods

We used linear models to determine the predictive relationship bnetween specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We chose the simplest model that explained the most variation in plant mass using Akaike's Information Criterion (AIC). We evaluated model assumptions for the simpolest model chosen using AIC forof normally distributed errors (residuals) using QQ plots and homoscedastic erros (errors) using Residuals vs Fitted and Scale-Locations plots. 

#### Goal of analysis

We used linear models to determine the predictive relationship between specific leaf area (cm2/g), treatment, and species on plant mass.

#### Model selection

We chose the simplest model that explained the most variation in plant mass using Akaike’s Information Criterion (AIC).

#### Model diagnostics

We evaluated model assumptions for the simplest model chose using AIC for normally distributed errors (residuals) using QQ plots and homoscedastic errors (residuals) using Residuals vs Fitted and Scale-Location plots.

### Results

We found that the simplest model decribing variation and plant mass included species and treatment (linear model: F(7, 62) = 27.08, p < 0.001, $\alpha$ = 0.05, adjusted R^2^ = 0.73). 







